---
title: "Datathon"
author: "Xinhui Jing"
date: "2025-10-01"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}


# -------------------------
# Load packages
# -------------------------
library(dplyr)
library(tsibble)
library(fable)
library(fabletools)
library(ggplot2)
library(lubridate)
library(tidyr)

```


```{r}
# ---- 1) Load data ----
df <- read.csv("dynamic_supply_chain_logistics_dataset.csv", stringsAsFactors = FALSE)


```

```{r}

# Convert timestamp -> Date (daily)
df$timestamp <- as.POSIXct(df$timestamp, tz = "UTC")
df$date <- as.Date(df$timestamp)

# Quick check
cat("Dataset shape:", dim(df)[1], "rows x", dim(df)[2], "cols\n")
cat("Columns available:\n")
print(colnames(df))

# Ensure required columns exist
reqs <- c(group_cols, metric, "date")
missing_cols <- setdiff(reqs, colnames(df))
if(length(missing_cols)>0) stop(paste("Missing required columns:", paste(missing_cols, collapse=", ")))
```



```{r}

# -------------------------
# 1️⃣ Prepare dataset
# -------------------------
df <- df %>% filter(!is.na(timestamp))
df$timestamp <- as.POSIXct(df$timestamp, format="%Y-%m-%d %H:%M:%S")
df <- df %>% mutate(year = year(timestamp))

# -------------------------
# 2️⃣ Create synthetic hierarchy
# -------------------------
df <- df %>%
  mutate(
    route_risk_bin = cut(route_risk_level, breaks = 6, labels = paste0("R", 1:6)),
    port_cong_bin  = cut(port_congestion_level, breaks = 6, labels = paste0("P", 1:6))
  )

# Ensure all combinations exist for all timestamps
all_keys <- expand.grid(
  timestamp = unique(df$timestamp),
  route_risk_bin = paste0("R", 1:6),
  port_cong_bin  = paste0("P", 1:6)
)

df_full <- all_keys %>%
  left_join(df, by = c("timestamp", "route_risk_bin", "port_cong_bin")) %>%
  mutate(disruption_likelihood_score = ifelse(is.na(disruption_likelihood_score), 0, disruption_likelihood_score)) %>%
  as_tsibble(key = c(route_risk_bin, port_cong_bin), index = timestamp)

# Build hierarchical tsibble
supply_hier <- df_full %>%
  aggregate_key(route_risk_bin / port_cong_bin,
                disruption_likelihood_score = sum(disruption_likelihood_score))

# -------------------------
# 3️⃣ Split train/test
# -------------------------
train_data <- supply_hier %>% filter(year(timestamp) %in% c(2021, 2022, 2023))
test_data  <- supply_hier %>% filter(year(timestamp) == 2024)

# Forecast horizon
n_test <- test_data %>%
  group_by(route_risk_bin, port_cong_bin) %>%
  summarise(n = n(), .groups = "drop") %>% pull(n) %>% max()

# -------------------------
# 4️⃣ Fill gaps for ETS
# -------------------------
train_data <- train_data %>%
  fill_gaps(.full = TRUE, disruption_likelihood_score = 0)

# -------------------------
# 5️⃣ Fit base ETS model
# -------------------------
base_fit <- train_data %>%
  model(ETS = ETS(disruption_likelihood_score))

# -------------------------
# 6️⃣ Forecast reconciliation
# -------------------------
recon_fc <- base_fit %>%
  reconcile(
    BU  = bottom_up(ETS),
    OLS = min_trace(ETS, method = "ols"),
    WLS = min_trace(ETS, method = "wls_struct")
  ) %>%
  forecast(h = n_test)

# -------------------------
# 7️⃣ Align forecast & test data for SMAPE
# -------------------------
fc_tbl <- recon_fc %>% as_tibble() %>%
  select(timestamp, route_risk_bin, port_cong_bin, .model, .mean)

smape_tbl <- fc_tbl %>%
  left_join(
    test_data %>% as_tibble() %>% select(timestamp, route_risk_bin, port_cong_bin, disruption_likelihood_score),
    by = c("timestamp", "route_risk_bin", "port_cong_bin")
  ) %>%
  rename(actual = disruption_likelihood_score, forecast = .mean)

# -------------------------
# 8️⃣ Compute accuracy metrics
# -------------------------
# Standard metrics
acc <- accuracy(recon_fc, test_data) %>%
  group_by(.model) %>%
  summarise(
    MAPE = mean(MAPE, na.rm = TRUE),
    RMSE = mean(RMSE, na.rm = TRUE)
  )

# SMAPE
smape_summary <- smape_tbl %>%
  group_by(.model) %>%
  summarise(
    SMAPE = mean(2 * abs(forecast - actual) / (abs(forecast) + abs(actual)), na.rm = TRUE)
  )

# Combine
acc_summary <- acc %>%
  left_join(smape_summary, by = ".model") %>%
  arrange(MAPE)

print(acc_summary)

# -------------------------
# 9️⃣ Optional: plot example for one route_risk_bin
# -------------------------
example_hist <- supply_hier %>%
  filter(route_risk_bin == "R3", is_aggregated(port_cong_bin))

example_fc <- recon_fc %>%
  filter(route_risk_bin == "R3", is_aggregated(port_cong_bin))

autoplot(example_fc, level = NULL) +
  autolayer(example_hist, disruption_likelihood_score, alpha = 0.3) +
  labs(title = "Reconciled forecasts – Route Risk R3 (total)",
       y = "Disruption Likelihood Score") +
  theme_bw()







# -------------------------
# 🔍 Coherence check table
# -------------------------
library(dplyr)

# Convert reconciled forecasts to tibble
fc_tbl <- recon_fc %>% as_tibble()

# Compute total forecast at top level
total_fc <- fc_tbl %>%
  filter(is_aggregated(route_risk_bin), is_aggregated(port_cong_bin)) %>%
  group_by(.model) %>%
  summarise(total = sum(.mean, na.rm = TRUE), .groups = "drop")

# Compute sum of bottom-level forecasts
bottom_fc <- fc_tbl %>%
  filter(!is_aggregated(route_risk_bin), !is_aggregated(port_cong_bin)) %>%
  group_by(.model) %>%
  summarise(sum_bottom = sum(.mean, na.rm = TRUE), .groups = "drop")

# Combine and calculate difference
coherence_tbl <- total_fc %>%
  left_join(bottom_fc, by = ".model") %>%
  mutate(diff_total_minus_bottom = total - sum_bottom)

print(coherence_tbl)


```


### What I observe in the chart:

Y-axis = Disruption Likelihood Score (0 to 1):

Most of the historical data points are very close to 1, which suggests that historically, disruptions were highly likely across Route R3.

Occasionally, you see sharp dips below 1, but they are rare.

Forecasts (coloured points by model: BU, ETS, OLS, WLS):

Near the end of the timeline (2024), your reconciled forecasts (BU, ETS, OLS, WLS) are plotted.

Notice that one of the models (purple, WLS) is forecasting a very low disruption score compared to the past.

This means some models predict conditions will improve, while others might remain conservative.

Historical pattern vs. forecast:

The historical trend shows persistent disruption risks.

Forecasting gives you an early warning system: which model predicts sustained risk, and which predicts reduction.



```{r}
```

